Large language models can propose mutations to any text-based artifact---prompts,
code, configurations---but using them as optimization oracles demands more than
iterative generation and manual inspection.
%
Existing approaches lack statistical rigor when comparing stochastic outputs,
default to greedy search with no mechanism to escape local optima, accumulate no
structured knowledge across experiments, and provide few safety guarantees around
cost or scope.
%
We present \textsc{anneal}, a domain-agnostic framework that treats LLM-guided
artifact optimization as a principled search problem.
%
\textsc{anneal} isolates every experiment in a git worktree, compares candidates
using the Wilcoxon signed-rank test with Holm--Bonferroni correction for
multiplicity control, and selects among search strategies---greedy, simulated
annealing, population-based, and Thompson Sampling---via a contextual bandit
meta-strategy.
%
A structured two-phase mutation pipeline first diagnoses the weakest evaluation
criteria, then generates targeted edits; episodic memory and a cross-target
learning pool transfer knowledge across experiments and domains.
%
Position-bias-cancelled Bradley--Terry scoring calibrates the LLM judge, while
adaptive sample sizing and held-out divergence detection control evaluation cost
and reliability.
%
On five optimization targets spanning prompt engineering, code optimization,
and documentation generation, \textsc{anneal} achieves up to 295$\times$
speedup on deterministic targets and consistent improvement on stochastic
targets within 50 experiments.
%
To our knowledge, \textsc{anneal} is the first open-source framework to unify
statistical hypothesis testing, structured metaheuristic search, and episodic
knowledge transfer for general-purpose LLM-guided artifact optimization.
