% !TEX root = ../main.tex
\section{Conclusion}
\label{sec:conclusion}

We presented Anneal, a domain-agnostic autonomous optimization framework that generalizes the mutate--evaluate--decide loop beyond its origins in ML training to arbitrary text-based artifacts.
The system operates on the minimal abstraction of an (Artifact, Eval, Agent) triplet: given any measurable artifact and an evaluation function, an LLM agent generates informed mutations, the engine evaluates them under statistical rigor, and a search strategy decides whether to accept or revert each change.
Every experiment is git-committed, scope-enforced, and logged.

The key contributions are:

\begin{enumerate}[nosep]
  \item A \textbf{stochastic evaluation framework} ($N$ samples $\times$ $K$ binary criteria) that extends autonomous optimization to domains with nondeterministic outputs---prompts, content, and creative artifacts---where deterministic scalar metrics are unavailable.

  \item A \textbf{knowledge compounding architecture} with TF-IDF retrieval, episodic memory, and cross-experiment learning transfer that enables the agent to accumulate domain-specific insights across hundreds of experiments, avoiding redundant exploration and enabling knowledge reuse across targets and projects.

  \item A \textbf{statistical decision framework} combining Wilcoxon signed-rank tests, Holm--Bonferroni correction, and bootstrap confidence intervals that reduces false positive acceptance by \NUM{fp-reduction}\% compared to uncorrected thresholding, ensuring that kept mutations represent genuine improvements.

  \item A \textbf{multi-timescale meta-optimization} system---two-phase mutation for per-experiment targeting, policy agent for continuous instruction adaptation, and component-level strategy evolution for structural search adaptation---that breaks through optimization plateaus where single-strategy approaches stagnate.

  \item An \textbf{open-source implementation} with cost controls, safety bounds, and operational tooling (live dashboard, webhook notifications, daemon mode) sufficient for unattended overnight optimization runs.
\end{enumerate}

Across five benchmark targets spanning prompt optimization, code optimization, and documentation generation, the full system achieves a \NUM{treat-vs-raw}\% mean improvement over sample-and-select baselines, with the iterative loop accounting for \NUM{greedy-vs-raw}\% and the enhancement suite contributing an additional \NUM{treat-vs-greedy}\%.
Treatment improvements are statistically significant on \NUM{sig-count} of 5 targets after Holm--Bonferroni correction, with medium-to-large effect sizes (Cohen's $d > 0.5$) on all targets.

Anneal is available under the Apache-2.0 license at \url{https://github.com/Mathews-Tom/anneal} and as the \texttt{anneal-cli} package on PyPI.
