% !TEX root = ../main.tex
\section{Experimental Evaluation}
\label{sec:evaluation}

We evaluate Anneal across three domains---prompt optimization, code optimization, and documentation generation---using a suite of five benchmark targets that exercise both evaluation modes (deterministic and stochastic) and a range of optimization objectives.

\subsection{Benchmark Suite}
\label{sec:benchmark-suite}

\cref{tab:benchmark-targets} summarizes the five benchmark targets.
Each target defines an artifact, an evaluation function, and a set of editable files;
the evaluation function and its criteria are immutable and outside the agent's mutation scope.

\begin{table}[t]
  \centering
  \caption{Benchmark targets. \emph{Eval mode}: D = deterministic (single scalar), S = stochastic ($N$ samples $\times$ $K$ binary criteria). \emph{Domain}: the optimization domain exercised by each target.}
  \label{tab:benchmark-targets}
  \small
  \begin{tabular}{@{}llllrl@{}}
    \toprule
    ID & Target & Domain & Eval & $K$ & Objective \\
    \midrule
    B1 & Classification prompt & Prompt opt. & S & 5 & Maximize classification accuracy \\
    B2 & Summarization prompt & Prompt opt. & S & 4 & Maximize summary quality \\
    B3 & Performance optimization & Code opt. & D & --- & Minimize execution time \\
    B4 & Code quality refactor & Code opt. & S & 6 & Maximize lint/style score \\
    B5 & README generation & Documentation & S & 5 & Maximize documentation quality \\
    \bottomrule
  \end{tabular}
\end{table}

The targets were selected to satisfy three coverage criteria.
First, they span both evaluation modes: B3 uses deterministic evaluation (a single scalar latency measurement), while B1, B2, B4, and B5 use the $N \times K$ stochastic evaluation framework with binary criteria.
Second, they cover three distinct optimization domains---prompt engineering, code transformation, and technical writing---ensuring that results are not domain-specific.
Third, they exercise different optimization objectives: maximization (B1, B2, B4, B5) and minimization (B3), testing the engine's directional generality.

Each target ships with a TOML configuration specifying artifact paths, scope constraints, evaluation commands, and scoring criteria.
The evaluation functions are self-contained: B3 runs a subprocess and parses wall-clock time; B1, B2, B4, and B5 invoke an LLM judge with fixed rubric criteria and aggregate binary pass/fail scores across $N = 10$ samples.

\subsection{Experimental Configurations}
\label{sec:configurations}

We compare four configurations of increasing capability, isolating the contribution of each enhancement layer:

\paragraph{Raw baseline.}
Generate $n = 50$ independent artifacts using the LLM (same model, same prompt template), evaluate each, and report the best score.
No optimization loop, no feedback, no knowledge accumulation.
This baseline measures pure sample-and-select performance without iterative refinement.

\paragraph{Greedy baseline.}
Run the Anneal engine with greedy search only: accept a mutation if and only if it produces a statistically significant improvement (Wilcoxon signed-rank test, $\alpha = 0.05$).
No knowledge injection, no strategy adaptation, no meta-optimization.
This isolates the contribution of the iterative mutate--evaluate--decide loop itself.

\paragraph{Control.}
Anneal with the core feature set: greedy search with Holm--Bonferroni correction, \texttt{program.md} mutation instructions, basic knowledge store (experiment history retrieval via TF-IDF similarity), and per-criterion structured feedback.
This configuration represents the engine without the advanced enhancements described in \cref{sec:method}.

\paragraph{Treatment.}
Anneal with all enhancements enabled: strategy manifest with component-level evolution, two-phase mutation (structured diagnosis before targeted edit), policy agent for continuous instruction rewriting, episodic memory with transferable lessons, simulated annealing search with adaptive reheat, multi-draft mutation with verifier pruning, and the research operator for external knowledge injection on plateaus.
This is the full system as described in \cref{sec:method}.

The progression from raw to treatment forms a nested ablation: each configuration adds a layer of capability, so pairwise comparisons isolate the marginal contribution of each layer.

\subsection{Statistical Design}
\label{sec:stat-design}

Each configuration runs 10 independent trials per benchmark target, each with a distinct random seed controlling LLM sampling temperature jitter and stochastic evaluation order.
Every trial executes 50 experiments (mutations), yielding a total experimental budget of:
\[
  5 \text{ targets} \times 4 \text{ configs} \times 10 \text{ runs} \times 50 \text{ experiments} = 10{,}000 \text{ experiments.}
\]

\paragraph{Pairwise comparison.}
For each pair of configurations on each target, we apply the Wilcoxon signed-rank test across the 10 paired final scores (paired by random seed).
This nonparametric test is appropriate because score distributions are not assumed normal, and pairing by seed controls for LLM sampling variability.

\paragraph{Multiple comparison correction.}
We apply Holm--Bonferroni correction across the five targets for each configuration pair, controlling the family-wise error rate at $\alpha = 0.05$.
This mirrors the correction used within the engine itself for consolidation-window decisions (\cref{sec:statistical-rigor}).

\paragraph{Effect size.}
We report Cohen's $d$ with 95\% bootstrap confidence intervals (10{,}000 resamples, bias-corrected and accelerated).
Bootstrap resampling uses deterministic seeding with float-precision normalization for cross-platform reproducibility.

\paragraph{Cost-normalized comparison.}
Because the treatment configuration incurs higher per-experiment cost (two-phase mutation, multi-draft generation), we additionally compare score-at-equivalent-budget: the best score each configuration achieves when constrained to the same total USD cost, rather than the same number of experiments.

\subsection{Metrics}
\label{sec:metrics}

\paragraph{Primary metrics.}
\begin{itemize}[nosep]
  \item \textbf{Final score}: the best score achieved across all 50 experiments in a trial.
  \item \textbf{Score at budget}: the best score achieved within a fixed cost budget (\$2.00 per trial), enabling cost-normalized comparison.
  \item \textbf{Convergence speed}: the experiment number at which 90\% of the final improvement (over the initial score) is first achieved.
  \item \textbf{Total cost}: cumulative USD cost per trial, computed from per-token pricing.
\end{itemize}

\paragraph{Secondary metrics.}
\begin{itemize}[nosep]
  \item \textbf{Mutation acceptance rate}: fraction of experiments where the mutation was kept (higher is not necessarily better; an acceptance rate near 1.0 suggests the evaluation is not discriminating).
  \item \textbf{Eval cache hit rate}: fraction of evaluations served from the content-hash cache, indicating redundant mutation generation.
\end{itemize}
