\section{Introduction}
\label{sec:introduction}

Large language models have demonstrated a striking capacity to act as
optimization oracles: given a text artifact and a description of what ``better''
means, an LLM can propose targeted mutations that improve performance on
measurable objectives~\cite{novikov2025alphaevolve, romera2024funsearch}.
This capability extends far beyond code---prompts, configuration files, system
instructions, and any other text-based artifact amenable to automated evaluation
are candidates for LLM-guided search.
Yet the gap between ``an LLM can suggest edits'' and ``an LLM reliably improves
artifacts'' remains wide.
Bridging it requires solving four interrelated problems that current tools
address only partially, if at all.

\paragraph{Statistical rigor.}
When evaluation is stochastic---an LLM judge scoring prompt outputs, a noisy
benchmark, a test suite with non-deterministic ordering---raw score comparisons
are unreliable.
A candidate that scores 0.82 versus a baseline of 0.80 may reflect genuine
improvement or sampling noise.
Prompt optimization tools such as DSPy~\cite{khattab2024dspy} and
EvoPrompt~\cite{guo2023evoprompt} typically compare mean scores without
hypothesis testing, leading to false-positive acceptances that compound over
sequential experiments.
The machine learning evaluation literature has long warned against this
practice~\cite{dror2018hitchhiker}, yet optimization frameworks rarely
incorporate the recommended statistical machinery.

\paragraph{Structured search.}
Most LLM-guided optimization systems use greedy hill-climbing: accept an edit if
the score improves, reject otherwise.
Greedy search is prone to local optima, particularly in the combinatorial space
of text mutations where small edits can interact non-linearly.
Metaheuristic search---simulated annealing, population-based methods, bandit
algorithms---offers well-understood mechanisms for escaping local optima and
balancing exploration with exploitation, yet these strategies have seen limited
adoption in LLM-guided settings.
AlphaEvolve~\cite{novikov2025alphaevolve} uses an island-based evolutionary
approach but is tightly coupled to code optimization;
Promptbreeder~\cite{fernando2024promptbreeder} evolves mutation operators but
provides no formal search strategy selection.

\paragraph{Knowledge accumulation.}
Each experiment in an optimization run generates information: what changed, what
improved, what regressed, and why.
Current systems discard this information after the accept/reject decision,
forcing the optimizer to rediscover lessons from prior experiments.
Effective optimization requires episodic memory---structured records of past
experiments that inform future mutations---and cross-target knowledge transfer
that lets insights from one optimization task accelerate another.

\paragraph{Safety and reproducibility.}
LLM-guided optimization involves executing generated code, running evaluation
commands, and making irreversible file changes.
Without isolation, a bad mutation can corrupt the working directory; without
budget controls, a stochastic evaluation loop can exhaust API credits overnight.
Reproducibility demands that every experiment state be recoverable, every
comparison be auditable, and every cost be tracked.

\medskip

No existing system addresses all four concerns simultaneously.
DSPy~\cite{khattab2024dspy} and OPRO~\cite{yang2024opro} focus on prompt
optimization without structured search or statistical testing.
AlphaEvolve~\cite{novikov2025alphaevolve} provides evolutionary search for code
but lacks statistical comparison and knowledge transfer.
ProTeGi~\cite{pryzant2023protegi} uses gradient-inspired textual feedback but
operates greedily with no memory.
FunSearch~\cite{romera2024funsearch} pairs LLMs with evaluators but is
restricted to function-level code search with best-shot selection.

\subsection{Contributions}

We present \textsc{anneal}, an open-source framework for autonomous artifact
optimization that addresses the four gaps above.
The framework makes three technical contributions---each validated as an
independent mechanism in \cref{sec:results}---supported by two architectural
contributions that realize the framework as a production-ready system.

\paragraph{Technical contributions (validated in \cref{sec:results}).}

\begin{enumerate}
  \item \textbf{Statistical acceptance under stochastic evaluation}
    (\cref{sec:gate-false-positives}).
    Stochastic evaluations are compared via the Wilcoxon signed-rank
    test~\cite{wilcoxon1945individual} with Holm--Bonferroni
    correction~\cite{holm1979simple} for sequential multiplicity.
    Position-bias-cancelled Bradley--Terry scoring~\cite{bradley1952rank}
    calibrates the LLM judge, and adaptive sample sizing extends or
    early-stops evaluation based on the observed effect size.
    On a synthetic null distribution, Holm--Bonferroni correction reduces
    false-positive acceptance by 86\% relative to uncorrected thresholding.

  \item \textbf{Adaptive metaheuristic search}
    (\cref{sec:gate-sa}).
    Beyond greedy hill-climbing, \textsc{anneal} implements simulated
    annealing with adaptive cooling and reheat, population-based search with
    tournament selection and LLM-guided crossover, island migration for
    diversity maintenance, and Pareto search for multi-objective optimization.
    A Thompson Sampling meta-strategy~\cite{thompson1933likelihood,
    chapelle2011empirical} selects among these algorithms based on observed
    reward, treating strategy selection itself as a bandit problem.
    On the 3D Rastrigin function, adaptive reheat yields a 19.1\% improvement
    over fixed geometric cooling.

  \item \textbf{Structured episodic knowledge retrieval}
    (\cref{sec:gate-retrieval}).
    Each experiment produces a structured \texttt{Lesson} capturing what
    changed, which criteria improved or regressed, and transferable insights.
    A cross-target learning pool shares knowledge across optimization tasks
    with domain-aware relevance filtering, a strategy manifest evolves the
    weakest component of the mutation approach when progress stalls, and a
    research operator injects external knowledge during plateaus.
    On a controlled similarity corpus, TF-IDF retrieval achieves
    $1.55\times$ the precision of Jaccard similarity.
\end{enumerate}

\paragraph{Architectural contributions (argued from design).}

\begin{enumerate}
  \setcounter{enumi}{3}
  \item \textbf{Git-native isolation and safety.}
    Every experiment runs in a dedicated git worktree, making all mutations
    reversible via \texttt{git reset}.
    Scope enforcement restricts agent edits to declared file paths, rejecting
    path traversal and absolute paths.
    Budget caps (per-experiment and daily), cost tracking per model, disk
    space checks, and process-group time-boxing prevent runaway resource
    consumption.

  \item \textbf{Domain-agnostic design.}
    The framework optimizes any text artifact evaluable by a shell command
    (deterministic mode) or an LLM judge against binary criteria (stochastic
    mode), with no domain-specific assumptions.
    Verification gates---binary pass/fail commands run before evaluation---let
    users enforce structural invariants (compilation, type-checking, format
    compliance) without spending evaluation budget on structurally invalid
    mutations.
\end{enumerate}

\subsection{Validation Approach}

We validate the three central mechanisms of \textsc{anneal}---statistical
acceptance, adaptive metaheuristic search, and structured knowledge
retrieval---on synthetic benchmarks with analytically known ground truth.
These benchmarks isolate each mechanism from the variance of a full
LLM-guided optimization run and allow direct quantitative comparison against
well-defined baselines: uncorrected sequential testing under a null
distribution, fixed-schedule simulated annealing on a nonconvex testbed, and
Jaccard set similarity on a controlled retrieval corpus.
\Cref{sec:results} reports the results and shows that each mechanism meets
its pre-registered gate threshold.
A full end-to-end evaluation of the complete framework against its baselines
across the five-target benchmark suite is left to future work, for the
reasons discussed in \cref{sec:empirical-scope,sec:threats}.

\subsection{Paper Organization}

\Cref{sec:related_work} surveys related work in LLM-guided optimization,
prompt engineering, metaheuristic search, and statistical evaluation.
\Cref{sec:system-design} describes the \textsc{anneal} architecture, including
the experiment loop, evaluation engine, search strategies, and knowledge system.
\Cref{sec:enhancements} details the enhancement tracks---representation, search,
knowledge, and evaluation---that build on the core loop.
\Cref{sec:evaluation} describes the benchmark suite and experimental design for
end-to-end configuration comparison.
\Cref{sec:results} presents mechanism-level validation results on synthetic
benchmarks.
\Cref{sec:discussion} discusses limitations and future work.
\Cref{sec:conclusion} concludes.
