\section{Related Work}
\label{sec:related_work}

The design space of LLM-guided artifact optimization admits a spectrum of
framework philosophies.
At one end are minimalist scaffolds that delegate almost all decision-making
to the coding agent itself: the framework provides an edit--evaluate--commit
loop, a single deterministic metric, and a human-maintained log of prior
attempts, and the agent's intrinsic reasoning supplies acceptance judgment,
search strategy, and memory of past experiments.
This design bet is that as frontier LLMs improve, elaborate scaffolding
becomes overhead rather than value, and that a thin framework plus a capable
agent is sufficient for optimization on problems with clean deterministic
metrics.

\textsc{anneal} occupies a different corner of this design space, one in
which the framework assumes responsibility for three decisions the agent
would otherwise make implicitly: whether a candidate is genuinely better
than its predecessor under stochastic evaluation, which search strategy
to use when the current one has stalled, and which prior experiments are
relevant to the current hypothesis.
These decisions are outside the reliable competence of a single-turn agent
call, particularly when the evaluation signal is noisy (LLM-judge scores,
small-sample benchmarks) and when the optimization trajectory spans dozens
of experiments.
The trade-off is more framework complexity in exchange for stronger
correctness guarantees and better use of accumulated experiment history.
The paper's three technical contributions (\cref{sec:results}) quantify
the value of each of the three decisions the framework makes on the
agent's behalf.

\cref{tab:related-comparison} summarizes the five closest prior systems
along the axes that distinguish \textsc{anneal}: the form of acceptance
decision, the search strategy layer, the knowledge accumulation
mechanism, the evaluation modes supported, and the scope of artifacts
optimized.
No prior system addresses all five axes simultaneously, though many cover
a subset.

\begin{table*}[t]
  \centering
  \caption{Comparison of \textsc{anneal} with the closest prior systems along
  five axes. \textbf{Accept:} how accept/reject decisions are made---point-score
  (P), best-shot selection (B), statistical test (S). \textbf{Search:} search
  strategy layer---greedy (G), evolutionary (E), bandit meta-strategy (M),
  reasoning-time search (R). \textbf{Memory:} knowledge accumulation
  mechanism---none (---), program database (D), verbal self-reflection (V),
  structured retrieval (S). \textbf{Eval modes:} deterministic (D),
  stochastic / LLM-judge (L). \textbf{Scope:} artifact type---code functions
  (C), prompts (P), reasoning traces (R), any text artifact (A).}
  \label{tab:related-comparison}
  \small
  \begin{tabular}{@{}llllll@{}}
    \toprule
    System & Accept & Search & Memory & Eval modes & Scope \\
    \midrule
    AlphaEvolve~\cite{novikov2025alphaevolve} & B & E & D & D     & C \\
    FunSearch~\cite{romera2024funsearch}      & B & E & D & D     & C \\
    DSPy~\cite{khattab2024dspy}               & P & G & --- & D,L   & P \\
    EvoPrompt~\cite{guo2023evoprompt}         & P & E & --- & D,L   & P \\
    Promptbreeder~\cite{fernando2024promptbreeder} & P & E & --- & D,L & P \\
    OPRO~\cite{yang2024opro}                  & P & G & --- & D,L   & P \\
    ProTeGi~\cite{pryzant2023protegi}         & P & G & --- & D,L   & P \\
    TextGrad~\cite{yuksekgonul2024textgrad}   & P & G & --- & D,L   & P \\
    Reflexion~\cite{shinn2023reflexion}       & P & G & V   & D,L   & R \\
    \midrule
    \textbf{\textsc{anneal}}                  & \textbf{S} & \textbf{G,E,M} & \textbf{S} & \textbf{D,L} & \textbf{A} \\
    \bottomrule
  \end{tabular}
\end{table*}

We organize prior work into five areas: LLM-guided code optimization,
LLM-guided prompt and reasoning optimization, evolutionary and metaheuristic
methods with LLMs, agent memory and self-reflection, and statistical rigor
in machine learning evaluation.
For each, we identify what the approach contributes and where \textsc{anneal}
differs or extends it.

\subsection{LLM-Guided Code Optimization}

The premise that LLMs trained on code can serve as plausible mutation
operators traces to early demonstrations that such models can produce
functionally correct edits from natural-language
instructions~\cite{chen2021codex}.
Recent work has built on this by coupling LLM proposal with automated
evaluators that close the loop from edit to signal.

\paragraph{AlphaEvolve.}
Novikov et al.~\cite{novikov2025alphaevolve} use an ensemble of LLMs within an
evolutionary framework to discover algorithms for mathematical and computational
problems.
AlphaEvolve operates on code functions, using an island-based population model
with a database of programs ranked by fitness.
The system achieved state-of-the-art results on problems ranging from matrix
multiplication to bin packing.
\textsc{anneal} shares the evolutionary search intuition---population-based
strategies and island migration are available search modes---but differs in three
respects.
First, \textsc{anneal} is domain-agnostic: it optimizes any text artifact, not
only code functions, by decoupling mutation from evaluation through a shell
command or LLM-judge interface.
Second, \textsc{anneal} applies statistical hypothesis testing (Wilcoxon
signed-rank with Holm--Bonferroni correction) rather than best-score selection,
which is critical when evaluation is stochastic.
Third, \textsc{anneal} accumulates structured knowledge across experiments via
episodic memory and cross-target learning pools, whereas AlphaEvolve relies on
the program database as implicit memory.

\paragraph{FunSearch.}
Romera-Paredes et al.~\cite{romera2024funsearch} pair an LLM with an automated
evaluator to search for novel mathematical functions, discovering new solutions
to the cap set problem and online bin packing.
FunSearch uses a best-shot selection strategy: the LLM generates many
candidates, and the evaluator retains only the top-scoring outputs.
This approach is effective when evaluation is cheap and deterministic but does
not extend to stochastic evaluation, where a single score comparison is
insufficient to distinguish signal from noise.
\textsc{anneal} supports the same deterministic pipeline---generate, evaluate,
keep the best---but adds paired statistical testing for stochastic settings,
structured search strategies beyond best-shot, and a knowledge system that
retains lessons from failed experiments rather than discarding them.

\paragraph{Eureka.}
Ma et al.~\cite{ma2023eureka} use LLMs as open-ended writers of reward
functions for reinforcement learning in simulated robotics, iterating on
the generated Python reward code against a deterministic task-success signal.
Eureka demonstrates that LLMs can propose targeted edits to executable
objective functions and that an outer loop over reward candidates improves
downstream RL agent performance.
The underlying loop---generate candidate code, evaluate via a deterministic
metric, keep the best---is structurally similar to the deterministic branch
of \textsc{anneal}'s evaluation pipeline; the difference is that Eureka
is tightly scoped to reward-function synthesis for RL, while \textsc{anneal}
is an artifact-agnostic framework whose evaluation function is user-defined.

\subsection{LLM-Guided Prompt and Reasoning Optimization}

Prompt optimization has generated a rich body of work, driven by the sensitivity
of LLM performance to prompt wording.
We review the most relevant systems.

\paragraph{Promptbreeder.}
Fernando et al.~\cite{fernando2024promptbreeder} evolve both task prompts and
the mutation operators that generate them, creating a self-referential
improvement loop.
Promptbreeder's key insight is that the space of mutation operators is itself
amenable to LLM-guided search.
\textsc{anneal} captures a related idea through its strategy manifest, which
decomposes the mutation approach into named components and evolves the weakest
component when progress stalls, and through the policy agent, which continuously
rewrites mutation instructions based on failure patterns.
Unlike Promptbreeder, \textsc{anneal} applies statistical testing to accept/reject
decisions and provides multiple search strategies rather than relying on a single
evolutionary scheme.

\paragraph{DSPy.}
Khattab et al.~\cite{khattab2024dspy} introduce a programming framework that
compiles declarative language model pipelines into optimized prompts and
fine-tuning data.
DSPy's optimizers (BootstrapFewShot, MIPRO) search over demonstrations and
instructions using teleprompter abstractions.
DSPy focuses on prompt pipeline compilation rather than general artifact
optimization, and its optimizers compare candidates by aggregate score without
hypothesis testing.
\textsc{anneal} addresses a broader scope---any text artifact with any evaluation
function---and uses the Wilcoxon test to control false-positive acceptance when
scores are noisy.

\paragraph{EvoPrompt.}
Guo et al.~\cite{guo2023evoprompt} apply discrete evolutionary operators
(genetic algorithm and differential evolution) to prompt optimization, treating
prompts as individuals in a population.
EvoPrompt demonstrates that evolutionary search outperforms greedy prompt
editing on several benchmarks.
\textsc{anneal} offers population-based search as one of several strategies and
adds Thompson Sampling~\cite{thompson1933likelihood} to adaptively select among
strategies, rather than committing to a single evolutionary algorithm.

\paragraph{OPRO.}
Yang et al.~\cite{yang2024opro} frame prompt optimization as an optimization
problem where an LLM generates new prompts given a history of (prompt, score)
pairs.
OPRO relies on the LLM's in-context learning to identify improving directions
from the optimization trajectory.
\textsc{anneal} similarly provides experiment history as context but structures
this history through episodic memory with per-criterion feedback, failure
taxonomy classification, and lineage tracking---rather than presenting raw
(prompt, score) tuples.

\paragraph{APE.}
Zhou et al.~\cite{zhou2023ape} automate prompt engineering by generating
candidate instructions, scoring them, and selecting the best.
APE uses a generate-and-filter approach without iterative refinement beyond
resampling.
\textsc{anneal} extends this with iterative mutation guided by structured
feedback from prior experiments and statistical testing for candidate comparison.

\paragraph{ProTeGi.}
Pryzant et al.~\cite{pryzant2023protegi} use textual gradients---natural
language descriptions of how to improve a prompt, generated by an LLM given
input-output error pairs---to guide prompt optimization.
The gradient metaphor provides directional feedback absent in random mutation.
\textsc{anneal}'s two-phase mutation pipeline serves a similar function: a
diagnosis step identifies the weakest evaluation criteria and their root causes
before a targeted mutation step generates edits, providing structured direction
without requiring the gradient abstraction.
ProTeGi operates greedily; \textsc{anneal} provides multiple search strategies
and statistical acceptance criteria.

\paragraph{TextGrad.}
Y\"uksekgonul et al.~\cite{yuksekgonul2024textgrad} formalize the textual
gradient metaphor as a differentiable-style computation graph in which LLMs
propagate natural-language feedback backward from a scalar loss through
composed components.
TextGrad extends ProTeGi's textual-gradient intuition to multi-component
pipelines, treating each component's feedback as an analogue of a gradient
update.
\textsc{anneal}'s diagnosis--mutation split is complementary: where TextGrad
generalizes gradient propagation through a computation graph, \textsc{anneal}
operates on a single artifact but adds statistical acceptance testing on top
of the resulting edits, ensuring that textual-gradient-style updates are not
accepted when they fall within evaluation noise.

\paragraph{Self-Refine.}
Madaan et al.~\cite{madaan2023selfrefine} introduce a canonical iterative
refinement loop in which a single LLM generates an output, critiques it, and
revises it without any external training signal, showing consistent
improvements across reasoning, code, and generation tasks.
Self-Refine establishes that iterative self-critique is valuable even when
the critic and generator are the same model.
\textsc{anneal} generalizes this loop in two directions: the critic is an
external evaluator rather than the LLM itself (removing the self-grading
bias), and the accept/reject decision is made by a statistical test on
paired evaluations rather than by the LLM's own judgment of whether a
revision is an improvement.

\paragraph{Tree of Thoughts.}
Yao et al.~\cite{yao2023tot} frame reasoning-time problem solving as
deliberate search over a tree of intermediate reasoning states, using
breadth-first or depth-first traversal with LLM-scored node evaluation.
Tree of Thoughts operates at reasoning time within a single model call,
producing a search over thought trajectories.
\textsc{anneal} operates at the outer loop instead---search over artifact
states across many model calls---but shares the intuition that structured
search over a model's proposals outperforms greedy single-shot generation
once the solution space has meaningful local optima.

\subsection{Evolutionary and Metaheuristic Methods with LLMs}

\paragraph{MAP-Elites.}
Mouret and Clune~\cite{mouret2015mapelites} introduce MAP-Elites, a
quality-diversity algorithm that maintains an archive of solutions spanning
a discretized behavioral space, keeping the highest-performing solution per
cell.
\textsc{anneal} incorporates a MAP-Elites archive as part of its knowledge
system, maintaining best solutions per behavioral region to preserve diversity
and enable the agent to draw on qualitatively different past solutions.

\paragraph{Thompson Sampling.}
Thompson Sampling~\cite{thompson1933likelihood} is a Bayesian approach to the
multi-armed bandit problem: sample from the posterior distribution of each arm's
reward and select the arm with the highest sample.
Chapelle and Li~\cite{chapelle2011empirical} provide an empirical evaluation
demonstrating its effectiveness relative to UCB and $\epsilon$-greedy
strategies.
\textsc{anneal} uses Thompson Sampling as a meta-strategy to select among search
algorithms (greedy, simulated annealing, population-based) based on each
algorithm's observed performance during the optimization run, treating strategy
selection as a contextual bandit problem with Beta-distributed priors updated
after each experiment.

\paragraph{Simulated annealing.}
Kirkpatrick et al.~\cite{kirkpatrick1983sa} introduced simulated annealing,
which accepts worse solutions with a probability that decreases as a temperature
parameter cools.
\textsc{anneal} implements simulated annealing with adaptive cooling and reheat:
when the acceptance ratio drops below a target threshold, the temperature
increases to restore exploration, preventing premature convergence.

\paragraph{Evolution through Large Models (ELM).}
Lehman et al.~\cite{lehman2023elm} propose using LLMs as the mutation operator
in a genetic algorithm, replacing hand-crafted mutation rules with
LLM-proposed edits to a seed program.
ELM establishes that LLM-as-mutation-operator is a viable alternative to
symbolic mutation in evolutionary program search.
\textsc{anneal} adopts the same proposal mechanism (LLM-generated edits to
a seed artifact) but places it inside a decision layer that applies
statistical acceptance rather than fitness-based selection, and generalizes
the target from programs to any text artifact.

\paragraph{LLaMEA.}
van Stein and B\"ack~\cite{vanstein2024llamea} present an LLM-driven
evolutionary algorithm for automatically generating metaheuristics,
applying LLM mutation operators to metaheuristic code and evaluating the
resulting algorithms on benchmark suites.
LLaMEA demonstrates that the LLM-as-mutation approach scales from
code-synthesis tasks to the generation of optimization algorithms
themselves.
\textsc{anneal} targets a different level: rather than generating new
metaheuristics, it composes existing metaheuristic strategies (greedy,
simulated annealing, population, Thompson Sampling meta-selection) into
a decision layer over LLM-generated artifact edits.

\subsection{Agent Memory and Self-Reflection}

A parallel line of work focuses on memory architectures that let LLM
agents accumulate knowledge across interactions, trials, or episodes.
These systems are relevant to \textsc{anneal}'s episodic memory and
cross-target learning pool.

\paragraph{Reflexion.}
Shinn et al.~\cite{shinn2023reflexion} equip LLM agents with verbal
self-reflection: after each trial, the agent produces a natural-language
critique of its own failure, which is appended to the context window for
subsequent trials.
Reflexion shows substantial gains on decision-making and reasoning
benchmarks from this lightweight memory mechanism.
\textsc{anneal}'s episodic memory carries a similar intuition---past
experiment outcomes inform future mutations---but extends it along two
axes: lessons are structured (\texttt{what\_changed},
\texttt{transferable\_insight}, \texttt{domain\_tags}) rather than
free-form critiques, and retrieval is similarity-based (TF-IDF or sentence
embeddings) rather than simply appending the full history.
This lets the agent see only the past experiments relevant to its current
hypothesis, rather than the entire trajectory.

\paragraph{Voyager.}
Wang et al.~\cite{wang2023voyager} build an open-ended Minecraft agent
whose central component is an incrementally-constructed skill library: as
the agent solves tasks, it abstracts working solutions into reusable
skills that are retrieved by semantic similarity for future tasks.
The skill library serves as a persistent memory that survives across
episodes.
\textsc{anneal}'s learning pool plays an analogous role for optimization
experiments: each kept mutation contributes a \texttt{Learning} to a
pool that is retrieved for new experiments on the same target or related
targets, with decay-adjusted impact and domain-aware relevance filtering.

\paragraph{MemGPT.}
Packer et al.~\cite{packer2023memgpt} introduce a hierarchical memory
architecture that treats the LLM's context window as managed working
memory with explicit paging to and from long-term storage.
MemGPT establishes that deliberate memory hierarchy---what to keep in
context, what to swap out, how to retrieve---is itself a system-design
problem rather than a property of the model.
\textsc{anneal} adopts a similar stance for experiment history: the
context-assembly layer decides which past experiments to inject into the
mutation agent's prompt based on recency, similarity, and a cold-start
policy, rather than naively including the full experiment log.

\subsection{Statistical Rigor in ML Evaluation}

\paragraph{Hypothesis testing for NLP.}
Dror et al.~\cite{dror2018hitchhiker} survey statistical significance testing
in NLP and recommend the Wilcoxon signed-rank test for paired comparisons of
system outputs, noting that parametric tests (paired $t$-test) assume normality
that NLP metrics rarely satisfy.
\textsc{anneal} implements this recommendation directly: stochastic evaluations
produce paired score vectors (challenger and baseline evaluated on the same
inputs), and the Wilcoxon signed-rank test~\cite{wilcoxon1945individual}
determines whether the observed differences are statistically significant.
When fewer than six paired samples are available, \textsc{anneal} falls back to
a Cohen's $d$ effect-size threshold, requiring at least a medium effect
($d > 0.5$) for acceptance.

\paragraph{Statistical power and variance in ML benchmarks.}
Card et al.~\cite{card2020power} analyze the statistical power of common
NLP benchmarks and find that many published comparisons are underpowered,
with effect sizes small enough that reported significant differences may
not replicate.
Bouthillier et al.~\cite{bouthillier2021variance} complement this analysis
by showing that variance in machine-learning benchmark results is
substantial and comes from multiple sources (data order, initialization,
hardware non-determinism), and propose that ML evaluation should report
paired variance estimates rather than single-run numbers.
These findings motivate \textsc{anneal}'s paired-seed evaluation protocol
and its use of variance-aware statistical tests at the accept/reject
boundary of every experiment, rather than only at a single final
comparison.

\paragraph{Multiple comparison correction.}
Holm~\cite{holm1979simple} proposes a sequentially rejective procedure that
controls the family-wise error rate without the conservatism of the Bonferroni
correction.
In \textsc{anneal}, each experiment within a consolidation window constitutes a
hypothesis test; the Holm--Bonferroni adjustment divides the significance level
$\alpha$ by the number of remaining comparisons in the window, reducing the
false-positive rate for sequential testing.
We validate the magnitude of this reduction on a synthetic null distribution in
\cref{sec:gate-false-positives}.

\paragraph{Bootstrap confidence intervals.}
Efron and Tibshirani~\cite{efron1994bootstrap} introduce the bootstrap as a
general-purpose resampling method for estimating the sampling distribution
of a statistic.
\textsc{anneal} uses the bias-corrected and accelerated (BCa) bootstrap with
10{,}000 resamples to produce confidence intervals on Cohen's $d$
(\cref{sec:stat-design}) and on aggregate stochastic-evaluation scores
(\cref{sec:eval-pipeline}), with deterministic seeding derived from the
score vector so that bootstrap results are reproducible across runs.

\paragraph{Bradley--Terry scoring and LLM-as-judge calibration.}
Bradley and Terry~\cite{bradley1952rank} propose a paired-comparison model where
each item has a latent strength parameter estimated from pairwise win/loss data.
\textsc{anneal} uses Bradley--Terry scoring with position-bias cancellation to
calibrate its LLM judge: when the number of evaluation votes is two or more,
half are presented with criteria in forward order and half in reverse, and the
resulting win rates are combined to estimate calibrated quality scores that
cancel the position bias inherent in LLM evaluation~\cite{zheng2023judging}.
This position-bias concern is well-established in recent LLM-as-judge
work: Liu et al.~\cite{liu2023geval} report that GPT-4 as a judge shows
measurable order effects and prompt sensitivity, and that calibration via
chain-of-thought and probability-weighted scoring yields better alignment
with human judgments than raw preference counts.
\textsc{anneal}'s Bradley--Terry aggregator is the mechanism-level
counterpart: rather than trusting raw vote counts, it computes a latent
strength parameter under the paired-comparison model, which is more
robust to the position-bias and binary-threshold artefacts that raw
counting inherits.
