% !TEX root = ../main.tex
\section{Mechanism Validation}
\label{sec:results}

A full end-to-end empirical comparison of the four configurations across all five benchmark targets is outside the scope of this paper: each configuration requires a multi-seed run on every target, and the associated compute and API budget is substantial.
Instead, we validate the three central mechanisms---statistical acceptance, adaptive metaheuristic search, and structured knowledge retrieval---in isolation on synthetic tasks where the ground truth is known analytically.
Each mechanism is evaluated against a specific quantitative gate defined before the experiment was run.
The gate benchmarks are implemented in \texttt{benchmarks/\allowbreak{}bench\_\allowbreak{}false\_\allowbreak{}positives.py}, \texttt{benchmarks/\allowbreak{}bench\_\allowbreak{}sa\_\allowbreak{}convergence.py}, and \texttt{benchmarks/\allowbreak{}bench\_\allowbreak{}retrieval\_\allowbreak{}precision.py}.
All results below are fully reproducible by invoking those scripts; each completes in under a minute and makes no LLM API calls.

% ==================================================================
\subsection{Statistical Acceptance under a Null Distribution}
\label{sec:gate-false-positives}

\paragraph{Setup.}
We construct a synthetic null distribution where no real improvement exists: paired score vectors of length $n = 10$ are drawn from $\mathcal{N}(0.5, 0.1)$ for both challenger and baseline on every experiment.
Any mutation accepted under this distribution is, by construction, a false positive.
We run 100 independent 50-experiment windows and count the average number of false positives per window under two decision rules:
(i) uncorrected comparison using a fixed significance threshold of $\alpha = 0.05$, and
(ii) Holm--Bonferroni-corrected comparison with the adjusted threshold $\alpha / (W - i)$ within a rolling window of size $W = 50$, as described in \cref{sec:stat-comparison}.

\paragraph{Result.}
Uncorrected comparison produces an average of 2.1 false positives per 50-experiment window.
Holm--Bonferroni correction reduces this to 0.3 false positives per window, a reduction of 86\%, comfortably exceeding the gate threshold of 50\% committed to in the benchmark harness before the experiment was run.
This result directly substantiates the statistical-rigor claim in \cref{sec:stat-comparison}: on stochastic evaluation signals, sequential acceptance without multiplicity control admits a non-trivial rate of spurious improvements, and the engine's correction machinery removes the majority of them while remaining valid under the standard family-wise error rate definition.

\begin{table}[t]
  \centering
  \caption{False positive rate under a synthetic null distribution ($\mathcal{N}(0.5, 0.1)$, $n = 10$ paired samples, $W = 50$-experiment windows, 100 replicates). The Holm--Bonferroni correction reduces false positives by 86\% with no loss of validity.}
  \label{tab:gate-fp}
  \small
  \begin{tabular}{@{}lr@{}}
    \toprule
    Decision rule & Avg. false positives / window \\
    \midrule
    Uncorrected ($\alpha = 0.05$) & 2.1 \\
    Holm--Bonferroni (rolling $W = 50$) & 0.3 \\
    \midrule
    Reduction & 86\% \\
    \bottomrule
  \end{tabular}
\end{table}

% ==================================================================
\subsection{Adaptive Simulated Annealing on Rastrigin}
\label{sec:gate-sa}

\paragraph{Setup.}
We evaluate the adaptive-reheat mechanism (\cref{sec:simulated-annealing}) against a fixed-temperature simulated annealing baseline on the 3-dimensional Rastrigin function, a standard nonconvex optimization testbed with a large number of local minima and a known global minimum at the origin.
Each configuration runs 50 independent trajectories of 100 experiments with matched seeds and identical starting temperatures ($T_0 = 10.0$, cooling rate $0.75$, minimum temperature $0.01$).
The fixed baseline disables reheat (\texttt{reheat\_factor} $= 1.0$); the adaptive configuration multiplies temperature by $2.0$ when the acceptance ratio over the last 10 experiments drops below half the target acceptance rate.
The primary metric is the best Rastrigin value encountered during the trajectory, since reheat is designed to escape local optima mid-run rather than to improve the asymptotic converged value.

\paragraph{Result.}
Fixed SA achieves an average best score of 18.3144 across the 50 trajectories.
Adaptive SA with reheat achieves 14.8228, a 19.1\% improvement.
Adaptive SA strictly outperforms fixed SA on 27 of the 50 seeds, with the remaining seeds either tied or marginally worse when the fixed schedule happens to cool into a basin containing the trajectory-best value before the reheat would fire.
The 19.1\% average improvement exceeds the 15\% gate threshold committed to in the benchmark harness before the experiment was run.

\begin{table}[t]
  \centering
  \caption{Adaptive vs. fixed simulated annealing on 3D Rastrigin ($T_0 = 10.0$, cooling rate $0.75$, 50 seeds $\times$ 100 experiments). Adaptive reheat yields a 19.1\% average reduction in the best Rastrigin value encountered, meeting the 15\% gate threshold.}
  \label{tab:gate-sa}
  \small
  \begin{tabular}{@{}lr@{}}
    \toprule
    Configuration & Avg. best Rastrigin score \\
    \midrule
    Fixed SA (no reheat) & 18.3144 \\
    Adaptive SA (with reheat) & 14.8228 \\
    \midrule
    Improvement & 19.1\% \\
    Adaptive wins / seeds & 27 / 50 \\
    \bottomrule
  \end{tabular}
\end{table}

This result substantiates the claim in \cref{sec:simulated-annealing} that adaptive reheat materially helps the search escape local optima in a multimodal landscape, relative to a fixed geometric cooling schedule.
The experiment uses a closed-form analytical objective rather than a learned LLM judge, isolating the contribution of the temperature schedule from any variance in the evaluation signal itself.

% ==================================================================
\subsection{Retrieval Precision for Episodic Memory}
\label{sec:gate-retrieval}

\paragraph{Setup.}
The knowledge system in \cref{sec:episodic-memory} retrieves past experiment records by similarity to the current hypothesis, upgrading from TF-IDF cosine similarity to sentence embeddings when available.
We evaluate the TF-IDF implementation against a Jaccard-similarity baseline on a controlled retrieval corpus.
The corpus consists of 30 synthetic experiment hypotheses organized into 5 groups of 6 hypotheses each, where intra-group hypotheses share a distinctive group-level keyword (e.g., ``gradient'', ``tensor'') against a background of shared generic terms.
The metric is precision@5: for each query, we retrieve the top 5 most similar documents and compute the fraction that belong to the same ground-truth group as the query.

\paragraph{Result.}
Jaccard retrieval achieves precision@5 of 0.647.
TF-IDF retrieval achieves precision@5 of 1.000, corresponding to a $1.55\times$ improvement ratio.
This exceeds the $1.5\times$ gate threshold committed to in the benchmark harness before the experiment was run.
The gap reflects the inverse-document-frequency weighting: shared generic terms inflate the Jaccard set union and dilute the discriminative signal, while TF-IDF down-weights those terms and surfaces the group-unique keyword.
In the context of episodic memory, this precision difference directly controls how often the system surfaces genuinely similar prior experiments---as opposed to superficially overlapping but topically unrelated records---when assembling agent context.

\begin{table}[t]
  \centering
  \caption{Retrieval precision@5 on a controlled similarity corpus (30 hypotheses, 5 ground-truth groups of 6). TF-IDF's IDF weighting yields a $1.55\times$ improvement over Jaccard by suppressing shared generic vocabulary.}
  \label{tab:gate-retrieval}
  \small
  \begin{tabular}{@{}lr@{}}
    \toprule
    Retrieval method & precision@5 \\
    \midrule
    Jaccard (word-level) & 0.647 \\
    TF-IDF (cosine) & 1.000 \\
    \midrule
    Ratio (TF-IDF / Jaccard) & $1.55\times$ \\
    \bottomrule
  \end{tabular}
\end{table}

% ==================================================================
\subsection{From Mechanism to System}
\label{sec:empirical-scope}

The three gate benchmarks above isolate the statistical, search, and knowledge subsystems and establish that each behaves as specified under controlled conditions.
They are not a substitute for an end-to-end evaluation of the full framework against its baselines across the benchmark suite described in \cref{sec:benchmark-suite}, which we treat as future work and discuss under threats to validity in \cref{sec:threats}.
The benchmark suite, its harness, and the analysis pipeline (\cref{sec:stat-design}) ship as part of the open-source distribution to enable independent replication at the scale the experimental design specifies.
